---
title: "Biostat 203B Homework 5"
subtitle: Due Mar 22 @ 11:59PM
author: "Yang An and UID: 106332601"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

## Predicting ICU duration

Using the ICU cohort `mimiciv_icu_cohort.rds` you built in Homework 4, develop at least three machine learning approaches (logistic regression with enet regularization, random forest, boosting, SVM, MLP, etc) plus a model stacking approach for predicting whether a patient's ICU stay will be longer than 2 days. You should use the `los_long` variable as the outcome. You algorithms can use patient demographic information (gender, age at ICU `intime`, marital status, race), ICU admission information (first care unit), the last lab measurements before the ICU stay, and first vital measurements during ICU stay as features. You are welcome to use any feature engineering techniques you think are appropriate; but make sure to not use features that are not available at an ICU stay's `intime`. For instance, `last_careunit` cannot be used in your algorithms. 

1. Data preprocessing and feature engineering.

2. Partition data into 50% training set and 50% test set. Stratify partitioning according to `los_long`. For grading purpose, sort the data by `subject_id`, `hadm_id`, and `stay_id` and use the seed `203` for the initial data split. 

```{r}
rm(list = ls())
library(tidymodels)
library(dplyr)
library(bigrquery)
library(dbplyr)
library(DBI)
library(gt)
library(gtsummary)
library(tidyverse)
set.seed(203)
# Load data
mimiciv_icu_cohort <- readRDS("/Users/yangan/Desktop/203B/203b-hw/hw4/mimiciv_shiny/mimic_icu_cohort.rds")

# sort
mimiciv_icu_cohort <- mimiciv_icu_cohort |>
  arrange(subject_id, hadm_id, stay_id) |>
  select(-admit_provider_id, -last_careunit, -hospital_expire_flag, -dod, 
         ) |>
  mutate(los_long = los > 2) |>
  mutate(age = anchor_age + year(intime) - anchor_year) |>
  mutate(age = ifelse(age < 0, NA, age)) |>
  mutate(age = as.numeric(age)) |>
  mutate(los_long = as.factor(los_long)) |>
  mutate(race = as.factor(race)) |>
  mutate(marital_status = as.factor(marital_status)) |>
  mutate(gender = as.factor(gender)) |>
  mutate(langeuage = as.factor(language)) |>
  mutate(insurance = as.factor(insurance)) |>
  mutate(admission_type = as.factor(admission_type)) |>
  mutate(admission_location = as.factor(admission_location)) |>
  mutate(discharge_location = as.factor(discharge_location)) |>
  mutate(age = as.numeric(age)) |>
  select(-los, -anchor_age, -anchor_year, -anchor_year_group) |>
  na.omit() |>
  print() 

data_split <- initial_split(
  mimiciv_icu_cohort, 
  # stratify by los_long
  strata = "los_long", 
  prop = 0.5
  )
```


```{r}
library(gtsummary)
# Numerical summaries stratified by the outcome `los_long`.
mimiciv_icu_cohort |> tbl_summary(by = los_long)
```


3. Train and tune the models using the training set.

```{r}
mimic_other <- training(data_split)
dim(mimic_other)
```
```{r}
mimic_test <- testing(data_split)
dim(mimic_test)
```
###svm

```{r}
svm_recipe <- 
  recipe(
    los_long ~ first_careunit + admission_location + insurance + language +
      marital_status + race + gender + age + wbc + bicarbonate + glucose + 
      sodium + creatinine + chloride + hematocrit + potassium + 
      `non invasive blood pressure systolic` + 
      `non invasive blood pressure diastolic` + 
      `respiratory rate` + `heart rate` + `temperature fahrenheit`,
    data = mimic_other
  ) |>
  # create traditional dummy variables (necessary for svm)
  step_dummy(all_nominal_predictors()) |>
  # zero-variance filter
  step_zv(all_numeric_predictors()) |> 
  # center and scale numeric data
  step_normalize(all_numeric_predictors()) |>
  print()

```

```{r}
svm_mod <- 
  svm_rbf(
    mode = "classification",
    cost = tune(),
    rbf_sigma = tune()
  ) |> 
  set_engine("kernlab")
svm_mod
```

```{r}
svm_wf <- workflow() |>
  add_recipe(svm_recipe) |>
  add_model(svm_mod)
svm_wf
```

```{r}
param_grid <- grid_regular(
  cost(range = c(-8, 5)),
  rbf_sigma(range = c(-5, -3)),
  levels = c(14, 5)
  )
param_grid
```

```{r}
set.seed(203)

folds <- vfold_cv(mimic_other, v = 5)
folds
```

```{r}
svm_fit <- svm_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
svm_fit
```




```{r}
svm_fit |>
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(x = cost, y = mean, color = factor(rbf_sigma))) +
  geom_point() +
  labs(x = "Cost", y = "CV AUC") +
  scale_x_log10()
```

```{r}
svm_fit |>
  show_best("roc_auc")
```
```{r}
best_svm <- svm_fit |>
  select_best("roc_auc")
best_svm
```

```{r}
# Final workflow
final_wf <- svm_wf |>
  finalize_workflow(best_svm)
final_wf
```

```{r}
# Fit the whole training set, then predict the test cases
final_fit <- 
  final_wf |>
  last_fit(data_split)
final_fit
```

```{r}
# Test metrics
final_fit |> 
  collect_metrics()
```
  
###random forest

```{r}
library(GGally)
library(gtsummary)
library(ranger)
library(tidyverse)
library(tidymodels)

rf_recipe <- 
  recipe(
        los_long ~ first_careunit + admission_location + insurance + language +
      marital_status + race + gender + age + wbc + bicarbonate + glucose + 
      sodium + creatinine + chloride + hematocrit + potassium + 
      `non invasive blood pressure systolic` + 
      `non invasive blood pressure diastolic` + 
      `respiratory rate` + `heart rate` + `temperature fahrenheit`,
    data = mimic_other
  ) |>
  # create traditional dummy variables (necessary for svm)
  #step_dummy(all_nominal_predictors()) |>
  # zero-variance filter
  step_zv(all_numeric_predictors()) |> 
  # center and scale numeric data
  #step_normalize(all_numeric_predictors()) |>
  # estimate the means and standard deviations
  # prep(training = Heart_other, retain = TRUE) |>
  print()
```

```{r}
rf_mod <- 
  rand_forest(
    mode = "classification",
    mtry = tune(),
    trees = tune()
  ) |> 
  set_engine("ranger")
rf_mod
```

```{r}
rf_wf <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_mod)
rf_wf
```

```{r}
param_grid2 <- grid_regular(
  trees(range = c(100L, 500L)),
  mtry(range = c(1L, 5L)),
  levels = c(5, 5)
  )
param_grid2
```

```{r}
set.seed(203)

folds <- vfold_cv(mimic_other, v = 5)
folds
```

```{r}
rf_fit <- rf_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid2,
    metrics = metric_set(roc_auc, accuracy)
    )
rf_fit
```




```{r}
rf_fit |>
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(x = trees, y = mean, color = factor(mtry))) +
  geom_point() +
  labs(x = "Num. of Trees", y = "CV AUC") 
```

```{r}
rf_fit |>
  show_best("roc_auc")
```
```{r}
best_rf <- rf_fit |>
  select_best("roc_auc")
best_rf
```

```{r}
# Final workflow
final_wf2 <- rf_wf |>
  finalize_workflow(best_rf)
final_wf2
```

```{r}
# Fit the whole training set, then predict the test cases
final_fit2 <- 
  final_wf |>
  last_fit(data_split)
final_fit2
```

```{r}
# Test metrics
final_fit2 |> 
  collect_metrics()
```

###XGBoost

```{r}
library(GGally)
library(gtsummary)
library(ranger)
library(tidyverse)
library(tidymodels)
library(xgboost)

gb_recipe <- 
  recipe(
        los_long ~ first_careunit + admission_location + insurance + language +
      marital_status + race + gender + age + wbc + bicarbonate + glucose + 
      sodium + creatinine + chloride + hematocrit + potassium + 
      `non invasive blood pressure systolic` + 
      `non invasive blood pressure diastolic` + 
      `respiratory rate` + `heart rate` + `temperature fahrenheit`,
    data = mimic_other
  ) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_date(all_datetime()) %>%
  step_rm(starts_with(c("intime", "outtime", "admittime", "dischtime", "deathtime", "edregtime", "edouttime", "dod")))
gb_recipe

```

```{r}
gb_mod <- 
  boost_tree(
    mode = "classification",
    trees = 1000,
    tree_depth = tune(),
    learn_rate = tune()
  ) |> 
  set_engine("xgboost")
gb_mod
```

```{r}
gb_wf <- workflow() |>
  add_recipe(gb_recipe) |>
  add_model(gb_mod)
gb_wf
```

```{r}
param_grid3 <- grid_regular(
  tree_depth(range = c(1L, 3L)),
  learn_rate(range = c(-5, 2), trans = log10_trans()),
  levels = c(3,10)
  )
param_grid3
```

```{r}
set.seed(203)

folds <- vfold_cv(mimic_other, v = 5)
folds
```

```{r}
gb_fit <- gb_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid3,
    metrics = metric_set(roc_auc, accuracy)
    )
gb_fit
```




```{r}
gb_fit |>
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(x = learn_rate, y = mean, color = factor(tree_depth))) +
  geom_point() +
  labs(x = "Learning Rate", y = "CV AUC") +
  scale_x_log10()
```

```{r}
gb_fit |>
  show_best("roc_auc")
```
```{r}
best_gb <- gb_fit |>
  select_best("roc_auc")
best_gb
```

```{r}
# Final workflow
final_wf3 <- gb_wf |>
  finalize_workflow(best_gb)
final_wf3
```

```{r}
# Fit the whole training set, then predict the test cases
final_fit3 <- 
  final_wf3 |>
  last_fit(data_split)
final_fit3
```

```{r}
# Test metrics
final_fit3 |> 
  collect_metrics()
```
### model stacking approach

```{r}
library(gtsummary)
library(keras)
library(ranger)
library(stacks)
library(tidyverse)
library(tidymodels)
library(xgboost)
mimic_recipe <- 
  recipe(
       los_long ~ first_careunit + admission_location + insurance + language +
      marital_status + race + gender + age + wbc + bicarbonate + glucose + 
      sodium + creatinine + chloride + hematocrit + potassium + 
      `non invasive blood pressure systolic` + 
      `non invasive blood pressure diastolic` + 
      `respiratory rate` + `heart rate` + `temperature fahrenheit`,
    data = mimic_other
  ) |>
  # create traditional dummy variables (necessary for xgboost)
  step_dummy(all_nominal_predictors()) |>
  # zero-variance filter
  step_zv(all_predictors()) 
  # estimate the means and standard deviations
  # prep(training = Heart_other, retain = TRUE)
mimic_recipe
```

```{r}
set.seed(203)
folds <- vfold_cv(mimic_other, v = 5)
folds
```
####logistic regression
```{r}
logit_mod <- 
  logistic_reg(
    penalty = tune(), 
    mixture = tune()
  ) |> 
  set_engine("glmnet", standardize = TRUE)
logit_mod
```

```{r}
logit_wf <- workflow() |>
  add_recipe(mimic_recipe) |>
  add_model(logit_mod)
logit_wf
```

```{r}
logit_grid <- grid_regular(
  penalty(range = c(-6, 3)), 
  mixture(),
  levels = c(100, 5)
  )

logit_res <- 
  tune_grid(
    object = logit_wf,
    resamples = folds,
    grid = logit_grid,
    control = control_stack_grid()
  )

logit_res
```
####random forest
```{r}
rf_mod2 <- 
  rand_forest(
    mode = "classification",
    # Number of predictors randomly sampled in each split
    mtry = tune(),
    # Number of trees in ensemble
    trees = tune()
  ) |>
  set_engine("ranger")
rf_mod2
```

```{r}
rf_wf2 <- workflow() |>
  add_recipe(mimic_recipe) |>
  add_model(rf_mod2)
rf_wf2
```

```{r}
rf_grid2 <- grid_regular(
  trees(range = c(100L, 500L)), 
  mtry(range = c(1L, 5L)),
  levels = c(5, 5)
  )

rf_res2 <- 
  tune_grid(
    object = rf_wf2, 
    resamples = folds, 
    grid = rf_grid2,
    control = control_stack_grid()
  )
rf_res2
```

####XGboost
```{r}
gb_mod1 <- 
  boost_tree(
    mode = "classification",
    trees = 1000,
    tree_depth = tune(),
    learn_rate = tune()
  ) |> 
  set_engine("xgboost")
gb_mod1
```

```{r}
gb_wf1 <- workflow() |>
  add_recipe(mimic_recipe) |>
  add_model(gb_mod1)
gb_wf1
```

```{r}
gb_grid1 <- grid_regular(
  tree_depth(range = c(1L, 3L)),
  learn_rate(range = c(-5, 2), trans = log10_trans()),
  levels = c(3,10)
  )

gb_res1 <- 
  tune_grid(
    object = gb_wf1, 
    resamples = folds, 
    grid = gb_grid1,
    control = control_stack_grid()
  )
gb_res1
```


```{r}
mimic_model_st <- 
  # initialize the stack
  stacks() |>
  # add candidate members
  add_candidates(logit_res) |>
  add_candidates(rf_res2) |>
  add_candidates(gb_res1) |>
  # determine how to combine their predictions
  blend_predictions(
    penalty = 10^(-6:2),
    metrics = c("roc_auc")
    ) |>
  # fit the candidates with nonzero stacking coefficients
  fit_members()
mimic_model_st
```

```{r}
mimic_model_st
```

```{r}
autoplot(mimic_model_st)
```

```{r}
autoplot(mimic_model_st, type = "members")
```

```{r}
autoplot(mimic_model_st, type = "weights")
```

```{r}
collect_parameters(mimic_model_st, "rf_res2")
```

```{r}
mimic_pred <- mimic_test %>%
  bind_cols(predict(mimic_model_st, ., type = "prob")) %>%
  print(width = Inf)
```

```{r}
yardstick::roc_auc(
  mimic_pred,
  truth = los_long,
  contains(".pred_FALSE")
  )
```
We can use the members argument to generate predictions from each of the ensemble members.
```{r}
mimic_pred <-
  mimic_test |>
  select(los_long) |>
  bind_cols(
    predict(
      mimic_model_st,
      mimic_test,
      type = "class",
      members = TRUE
      )
    ) |>
  print(width = Inf)
```

```{r}
map(
  colnames(mimic_pred),
  ~mean(mimic_pred$los_long == pull(mimic_pred, .x))
  ) |>
  set_names(colnames(mimic_pred)) |>
  as_tibble() |>
  pivot_longer(c(everything(), -los_long))
```

4. Compare model classification performance on the test set. Report both the area under ROC curve and accuracy for each machine learning algorithm and the model stacking. Interpret the results. What are the most important features in predicting long ICU stays? How do the models compare in terms of performance and interpretability?

